Management induced changes of soil organic carbon on global croplands

### 2020-12-14 SOCbudgetPaper/discussion.Rmd
update discussion 2

## Agricultural management matters

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt considerably compared to constant management assumptions. This result remains robust to variations of central model parameters and initialization stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to net reduction of the SOC debt.

The evaluation of SOC stock gains and losses, however, is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 the baseline and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see [@pugh_simulated_2015]), global croplands have acted as a carbon sink according to our study. Annual C sequesteration rates of 0.2 per mille are well below the promoted 4 per mille ([@minasny_soil_2017]), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

@@

According to [@sanderman_soil_2017] SOC debt since the beginning of human cropping activities have been at around 37 GtC for the first 30 cm of the soil with half of it attributed to grazing. Our estimate of 22 GtC in 2010 for cropland debt is higher as [@sanderman_soil_2017] estimations. However, there are large uncertainties in modeling SOC at the global scale, and the [@sanderman_soil_2017] pointed out that their results might be conservatively low compared to experimental results, suggesting that our results are in an acceptable range.

Furthermore, [@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to time-variant agricultural management. [@pugh_simulated_2015] considered management effects like tillage and residue returning in a static way, but neither changes over time nor alignment to observed historical data like yields-levels or no-tillage areas were taken into account. The study moreover concludes that crop productivity gains (increasing yield levels by 18%) do not lead to a substantial decline in SOC debt (less than 1% change).

@@

Moreover, our sensitivity analysis indicates that (1) higher returned residue biomass driver by yield increases play the most important role, followed by (2) greater manure applications due to an increasing livestock sector as well as improved animal waste management systems, and (3) the adoption of no tillage practices. Unfortunately, disentangling drivers of higher residue and manure inputs remains unsolved within this study.

@@

Modeling management effects at the global scale, however, comes with parametric and structural uncertainties. High SOC gains in the first 30 cm of the soil compared to natural vegetation in e.g. the temperate croplands of Central Europe and UK, indicate suspicious high estimations of SOC inputs. As pointed out by [@keel_large_2017] and [@smith_how_2020], carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. [@keel_large_2017] question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries, which might not be reflected in outdated data sources. Following this line of argument, SOC results calculated in this study may overestimate actual SOC stocks due to an overestimation in residue biomass, for which the ratio compared to the harvested organ is normally reduced due to breeding. However, with an evaluation of management effects (see Sect. \@ref(sec:ipcccompare)), there is no indication for a systematic overrating of residue biomass. Rather, it is also likely that we are still missing carbon inputs e.g. from cover- and inter-cropping to the soil.

Another aspect concerns the initialization of SOC stocks in 1901, which is assumed to be in steady state considering the land-use pattern of 1901 and agricultural management data of 1965. As shown in Fig. \@ref(fig:SOCinit) the SOC debt estimation halves (from ~26 GtC to ~14 GtC) as well as the decrease in SOC debt (from ~4 GtC to ~2.5 GtC), if considering initialization SOC stocks under undisturbed natural vegetation. [@pugh_simulated_2015] pointed out the importance of accounting for the land-use history, as many CO~2~ emissions from agricultural soils are caused by historic land-use change (LUC) and the slow decline of SOC under cropland before it reaches a new equilibrium. Our results show that these legacy fluxes are compensated by agricultural management changes even under the strong assumption of the Initial-natveg scenario. However, it remains open, if our simple, depth-restricted SOC model itself can represent these fluxes correctly.

Generally, the limit to the first 30 cm of the soil profile follows the IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]) and assumes, that most of the SOC dynamic happens in the topsoil. In this regard several aspects are over simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Secondly changes to the subsoil due to tillage are neglected, which might be important to noticed, hence studies ([@powlson_limited_2014]) have shown, the subsoil to be a game changer in evaluating total SOC losses or gains for no-tillage systems. Thirdly, emissions from croplands on organic soils (like peat- and wetlands) are likely underestimate substantially.

This study excludes not only peatland degradation, it also does not account for carbon displacement via leaching and erosion. However, as pointed out by [@doetterl_erosion_2016], the final fate of leached or eroded carbon is uncertain and might even offset LUC emissions ([@wang_human-induced_2017]). Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focusing especially on the SOC debt, displaced but not emitted SOC can be treated as SOC stayed on croplands.
Moreover, the exclusive focus on croplands ignores LUC emission on other land-use types such as pastures, rangelands and forestry. Human interventions have led to large changes in SOC stocks there as well ([@sanderman_soil_2017],[@friedlingstein_global_2019]). This study does not try to be an comprehensive LUC emission analysis and acknowledges that land-use changes comes with large overall emissions.

@@

To validate our modeled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]) which are based on measurement data for croplands (see Table \@ref(tab:SCFtable)). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones (Fig. \@ref(append:climatemap)).

@@

Whereas our estimates are too low in the two tropical climate zones, temperate zone default factors are overestimated for the refinement of the IPCC guidelines ([@calvo_buendia_ipcc_2019]). Admittedly, the tropical factor differ substantially between the IPCC guidelines in 2006 ([@eggleston_hs_ipcc_2006]) and their refinement in 2019 ([@calvo_buendia_ipcc_2019]), leaving our tropical results in between.
Taking the oversimplified assumption, that tropical soils might suffer from insufficient C input rates (low) due to yield gaps , whereas temperate soils in developed regions might be highly managed and fertilized (high), the difference vanish partially. Our results show still especially in temperate dry regions -- the smallest region area-wise -- small deviations from natural SOC stocks. Considering the impact of irrigation ad fertilization on carbon-poor dryland soils, even factors above 1 (see Fig. \@ref(fig:SOCmaps)(c)) might be not surprising.

@@

The global estimates of SOC stock from this study are on the lower end compared to other modeled results or data driven estimates. Looking at regional results in Fig. \@ref(append:regcompare), our estimates turn out to be in good agreement for most of the world, with the largest deviations for boreal regions. To avoid that this bias influences our results, our study focuses on SOC changes on cropland. Pristine natural vegetated areas without human land management cancel out in calculation of $\Delta SOC$ and $F^\mathrm{SCF}$.

Additionally, our estimates for total SOC stocks of the world, as well as our SOC initialization, are dominated by the representation of natural vegetation, which are, however, only estimated in a basic manner. For example, we do not differentiated parametrization of nitrogen and lignin content of litterfall for woody and grass plant types. This leaves carbon inputs and decay behavior for natural litterfall rather uncertain. The absolute values of SOC stocks and debt from land-use change have to be used with caution. As our default litter parameterization accounts for woody plant types, larger uncertainty in natural land may arise especially in less forested areas.

We conducted a sensitivity analysis (Fig. \@ref(append:senslitterpnv) based on various plant parameterizations from the Century model (see Sect. \@ref(sec:scen_litterpnv)). This shows that the general trend of decreasing SOC debt of ~4 GtC is not altered under various estimates for natural SOC stocks.

>>>>

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt considerably compared to the often used constant management assumptions. This result remains robust to variations of central model parameters and variations in the initialization of SOC stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to a net reduction of the SOC debt since 1975.

@@

The evaluation of SOC stock gains and losses, is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see [@pugh_simulated_2015]), global croplands have acted as a carbon sink since 1975 according to our study. However, annual C sequesteration rates of 0.2 per mille are well below the promoted 4 per mille ([@minasny_soil_2017]), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

According to [@sanderman_soil_2017], the SOC debt since the beginning of human cropping activities has been at around 37 GtC for the first 30 cm of the soil with half of it attributed to SOC depletion on grasslands. Our estimate of 22 GtC in 2010 for cropland debt is higher as [@sanderman_soil_2017] estimations. However, there are large uncertainties in modeling SOC at the global scale, and [@sanderman_soil_2017] pointed out that their results might be conservatively low compared to experimental results. This suggests that our results are within a plausible range.

Furthermore, [@sanderman_soil_2017] modeled historical trends based on agricultural land expansion without considering SOC variations due to time-variant agricultural management. [@pugh_simulated_2015] considered management effects like tillage and residue returning in a static way, but neither changes over time nor alignment to observed historical data like yield-levels or no-tillage areas were taken into account. Their study moreover concludes that crop productivity-gains (increasing yield levels by 18% in their simulations) do not lead to a substantial decline in SOC debt (less than 1% change). Historic productivity increases were, however, notably larger. Despite large spatial heterogeneity aggregate yield increase rates are often estimate well above 50% ([@pellegrini_crop_2018], [@ray_recent_2012], [@rudel_agricultural_2009]).

@@

Looking on historical cropland intensification rates of above 50% ([@rudel_agricultural_2009]), a large fraction of increased C input from residue biomass, is attributed to productivity improvements.

@@

Modeling management effects at the global scale comes with parametric and structural uncertainties. High SOC gains in the first 30 cm of the soil compared to natural vegetation in e.g. the temperate croplands of Central Europe and the United Kingdom indicate suspiciously high estimates of SOC inputs. As pointed out by [@keel_large_2017] and [@smith_how_2020], carbon input calculations are highly sensitive to the choice of allometric functions determining below and above ground residue estimates from harvested quantities. [@keel_large_2017] question whether below ground residues might increase with a fixed root:shoot ratio rather than being independent of productivity gains. Moreover, the study pointed out that plant breeding shifts allometries, which might not be reflected in outdated data sources. While our study considers a dynamic harvest index with rising yields for several crops, we may still overestimate residue biomass in particular for belowground biomass. However, looking on the evaluation of management effects (see Sect. \@ref(sec:ipcccompare)), there is no general indication of overestimating stock change factors at least compared to IPCC default assumptions. Moreover, it is also likely that we are still missing carbon inputs to the soil from e.g. cover- and inter-cropping practices.

Another uncertainty is connected with the initialization of SOC stocks in 1901, which is assumed to be in steady state considering the land-use pattern of 1901 and agricultural management data of 1965. As shown in Fig. \@ref(fig:SOCinit) the SOC debt estimate almost halves (from ~26 GtC to ~14 GtC), and the SOC debt reduction is strongly reduced (from ~4 GtC to ~2.5 GtC), if considering initialization SOC stocks under undisturbed natural vegetation. [@pugh_simulated_2015] pointed out the importance of accounting for the land-use history, as many CO~2~ emissions from agricultural soils are caused by historic land-use change (LUC) and the slow decline of SOC under cropland before it reaches a new equilibrium. Our results of the Initial-natveg scenario show that the qualitative finding of a reduction of SOC debt through improved agricultural management is robust to changing the initialisation of soil organic carbon, even though the level of SOC debt is sensitive to the initialization setting.

Generally, the limit to the first 30 cm of the soil profile follows the IPCC guidelines ([@eggleston_hs_ipcc_2006], [@calvo_buendia_ipcc_2019]) and assumes that most of the SOC dynamic happens in the topsoil. In this regard several aspects are strongly simplified within our approach. Firstly, distribution of carbon inputs into different soil layers are neglected and all carbon inputs are allocated to the topsoil. This particularly overestimates SOC stocks in the first 30 cm of soil below deeper rooting vegetation, which is certainly the case for most of the woody natural vegetated areas. Secondly changes to the subsoil due to tillage are neglected. As [@powlson_limited_2014] have shown, the subsoil can be a game changer in evaluating total SOC losses or gains for no-tillage systems. No-tillage effects may seem larger than the actually are, if only focusing on the topsoil. SOC transfer to deeper soil layers under tillage, might enhance subsoil SOC compared to no-till practices. Thirdly, organic soils (like peat- and wetlands) and drained cropland areas are not explicitly considered and emissions from these cropland areas are thus likely substantially underestimated.

This study excludes not only peatland degradation, it also does not account for carbon displacement via leaching and erosion. However, as pointed out by [@doetterl_erosion_2016], the final fate of leached or eroded carbon is uncertain and might even offset LUC emissions ([@wang_human-induced_2017]). Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focusing especially on the SOC debt, displaced but not emitted SOC can be treated as SOC that remains on the cropland.
Moreover, the exclusive focus on croplands ignores LUC emission on other land-use types such as pastures, rangelands and forestry. Human interventions have led to large changes in SOC stocks there as well ([@sanderman_soil_2017],[@friedlingstein_global_2019]). This study does not intend to be a comprehensive LUC emission analysis and acknowledges that land-use changes comes with large overall emissions.

@@

To validate our modeled SOC stocks and stock changes under management, we compare our results to default IPCC stock changes factors of 2006 ([@eggleston_hs_ipcc_2006] and their refinements in 2019 [@calvo_buendia_ipcc_2019]). Both estimates are based on measurement data for croplands (see Table \@ref(tab:SCFtable)). To allow for comparison, we aggregate our stock change factors to the four IPCC climate zones (Fig. \@ref(append:climatemap)).

@@

Whereas our estimates are lower in the two tropical climate zones, temperate zone default factors are higher than [@calvo_buendia_ipcc_2019]. The tropical factors differ substantially between the IPCC guidelines in 2006 ([@eggleston_hs_ipcc_2006]) and their refinement in 2019 ([@calvo_buendia_ipcc_2019]), with our estimates within this range.
Taking the simplified assumption, that tropical soils might suffer from insufficient C input rates (low) due to yield gaps, whereas temperate soils in developed regions might be highly managed and fertilized (high), the difference vanishes partially. Our results still show  especially in temperate dry regions -- the smallest region area-wise -- small deviations from natural SOC stocks. Considering the impact of irrigation and fertilization on carbon-poor dryland soils, even factors above 1 (see Fig. \@ref(fig:SOCmaps)(c)) may be expected.

@@

The global estimates of the total SOC stock from this study are on the lower end compared to other modeled results or data driven estimates. SoilGrids ([@hengl_soilgrids250m_2017]) especially stands out for their high estimation, whereas all other sources (including out study) are comparabely similar. Looking at regional results in Fig. \@ref(append:regcompare), our estimates turn out to be in good agreement for most of the world, with the largest deviations for boreal moist and tropical moist areas. To avoid that this bias influences our results, which originates from uncertainties in the representation of natural land, we focus on SOC changes on cropland. Pristine natural vegetated areas (like permafrost and rain forests) without human land management thus drop out in our calculation of $\Delta SOC$ and $F^\mathrm{SCF}$.

Additionally, our estimates for total SOC stocks of the world (as well as our SOC initialization) are dominated by the representation of natural vegetation, which are only estimated in a basic manner. For example, we do not differentiate the parameterization of nitrogen and lignin content of litterfall for woody and grass plant types. This renders carbon inputs and decay dynamics for natural litterfall rather uncertain. The absolute values of SOC stocks and debt from land-use change have to be interpreted with caution. As our default litter parameterization accounts for woody plant types, larger uncertainty in natural land SOC dynamics may arise especially in less forested areas.

We conducted a sensitivity analysis (Fig. \@ref(append:senslitterpnv) based on various plant parameterizations from the Century model (see Sect. \@ref(sec:scen_litterpnv)). This shows that the general trend of decreasing SOC debt of ~4 GtC within the period of 1975--2010 is not altered under various estimates for natural SOC stocks.
